Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders

### 2025-07-25 Paper/Paper.qmd
update paper

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
    data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "Taxometric methods are designed to test whether latent structures underlying psychological constructs are categorical (taxonic) or dimensional. Despite their relevance to the nature of neurodevelopmental disorders (NDDs), their application has been limited and often methodologically flawed. We conducted a preregistered systematic review to critically evaluate this literature. The review identified all published taxometric studies on autism, ADHD, learning, and language disorders, with a focus on methodological biases such as artificial admixture. While studies on ADHD and SLI supported dimensional structures, most studies on autism and NDD subtypes reported taxonic findings, often in the presence of methodological threats. A Monte Carlo simulation illustrates how artificial admixture can spuriously generate taxonic results in otherwise dimensional data. We conclude that evidence for categorical models in NDDs is frequently undermined by methodological artifacts, reinforcing the need for caution, but also advocating for more, high-quality taxometric studies in the field of NDDs."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

@@

In recent years, research on neurodevelopmental disorders (NDDs) has undergone significant reconsideration. Traditional diagnostic categories defined in the DSM-5 [@dsm5] have increasingly been questioned in favor of a dimensional and transdiagnostic perspective [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial]. Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder [ASD] in place of distinct diagnoses). Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent, at least in part, the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

Determining whether NDDs truly represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial]. But more fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.

A dimensional approach aligns well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial]. It also seems well-suited for NDDs, given a growing body of evidence supporting polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome]. Most behavioral traits are understood to result not from one or a few causes, but from the additive effects of many small influences [@chabris2015fourth]. The central limit theorem [@fischer2011central] posits that, when many individual causes (such as discrete genetic variants) contribute additively, the resulting distribution tends to continuity and normality (this is, in fact, the "secret" of the Gaussian distribution). Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the proposed emergence of data-driven taxa [@astle2022trandiagnostic].

Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues in the 1970s and refined over decades [e.g., @meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure. It has been widely applied in psychopathology, and typically suggested that most mental health conditions (e.g., depression, anxiety) are better conceptualized as dimensional [@haslam2012review; @haslam2020metaanalysis]. In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to compare competing latent models: one categorical, the other dimensional [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric]. These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide a quantitative measures to facilitate inference [@ruscio2018using].

Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, which most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis]. In their debate on whether ASD is best conceptualized as a category or a continuum, @happe2021both and @chown2021bit do not even mention such methods. Similarly, in their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but puts much more emphasis on studies using cluster analysis. However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues]. Recent simulations using psychologically realistic data (featuring both skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters]. Indeed, @astle2022trandiagnostic interpret clustering results as evidence that data-driven groups do not align with DSM-5 categories, rather than as compelling tests of latent structural form.

@@

Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition (i.e., the putative taxon and complement). Ideally, Cohen’s *d* for each indicator should substantially exceed 1 (*d* $\gg$ 1). Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership (indicating redundancy among variables). Low base rate means that too few individuals in the sample belong to the taxon, making accurate detection difficult (samples with *n* < 30 for the taxonic group are usually deemed insufficient). This is particularly relevant when studying rare diagnostic categories. Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate. For instance, clinical cases may be recruited separately from controls and merged into a single sample. In general, poor indicator validity and low base rate reduce power to detect true taxa, while high nuisance covariance and artificial admixture increase the risk of detecting pseudotaxa by mimicking a categorical structure.

@@

The primary aim of this systematic review is to critically examine the use of taxometric methods in studies of NDDs within the published literature. We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment. (Language disorders were not included in the preregistered protocol and therefore constitute a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.

A crucial goal was to assess the methodological quality and credibility of taxometric findings. We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and skewed indicator distributions. (The latter is particularly relevant when scales assess rare or extreme behaviors, which can create the illusion of a categorical structure.)

Following the review, we report a Monte Carlo simulation study designed to empirically demonstrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data, using both taxometric and clustering methods.

@@

Inclusion criteria were as follows:  
1. The study applied at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);  
2. The target population consisted of individuals with NDDs or related traits;  
3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) for an independent reader to infer such a conclusion.

Exclusion criteria were:  
1. The article was theoretical, methodological, or a simulation study without empirical data;  
2. The study did not report any behavioral data (e.g., only neuroimaging or neurophysiological data; note: this criterion was later reconsidered, but no studies were excluded for this reason in practice);  
3. The study was a review and/or meta-analysis;  
4. The study employed latent structure methods other than taxometric analysis (e.g., latent profile analysis, mixture modeling), with no taxometric procedures reported.

In addition, consistent with the preregistered protocol, the search was limited to published, peer-reviewed journal articles. This decision reflected the primary aim of providing a critical review of the existing empirical literature. However, one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Discussion* section.

@@

Two independent human coders (MC and RP) screened and coded all articles. No AI tools were involved in the coding process, except in occasional cases where coders used them to help clarify specific article content. All discrepancies were resolved through discussion and later reviewed by the principal investigator (ET).

@@

```{r}
#| label: tbl-summary
#| tbl-cap: Overview summary of the reviewed articles."
summary_tab = read_excel("Tables/Review-Summary.xlsx",sheet=2)
kable(summary_tab, format = "latex", booktabs = TRUE) |>
  kable_styling(font_size = 8, latex_options = c("striped", "scale_down")) |>
  column_spec(1:ncol(summary_tab),width="5em") |>
  landscape()
```

@@

# Simulation of Artificial Admixture

We focus on artificial admixture because it represents a powerful threat to validity that is both common (as suggested by our review) and avoidable by design through better planning of a study. Different types of admixture can be incurred. @ruscio2004checklist describes three main forms. First (and most frequently) is the case where clinical and nonclinical participants are separately recruited and merged into a single sample. Second, admixture can occur when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf. "subtractive" compound sampling in @haslam2020metaanalysis]. Third, researchers may trim observations from the putative complement (i.e., individuals without the disorder) to inflate the base rate of the taxon. Each of these procedures risks imposing a categorical structure onto data that might be dimensional in nature. It is worth noting that artificial admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the same reasons: both infer latent structure based on observed covariation, which can be distorted by selection artifacts.

@@

We simulate a dimensional scenario to visually demonstrate how artificial admixture, specifically of the first type described above, can produce pseudotaxonic results. A single normally distributed latent variable $X$ is generated to represent the underlying clinical trait of interest. A set of observed indicators ($x_1$, $x_2$, $x3$) are then simulated as linear functions of $X$ plus normally distributed random noise, yielding an $R^2$ of 0.77 for each indicator. Individuals above the 99th percentile of $X$ are assumed to receive a diagnosis via global clinical assessment. A sample of $N = 1000$ individuals is drawn from the general population. Given the low base rate, only about $n = 10$ cases are expected to naturally receive a diagnosis. To address this, researchers draw a second sample consisting of an additional $n = 1000$, targeting clinically diagnosed cases, and merge it with the original sample, thereby creating an artificial admixture. The full simulation code and data are available in the online materials:  
<https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba>

@@

Concerning subtypes, the situation was similar. All four reviewed articles reached some taxonic conclusions, but three of them likely involved some form of artificial admixture. However, the issue appeared to be nuanced and less severe in this subset of studies. For example, @obrien2012dyslexia used psychometric cutoffs across different tools and inclusion criteria, which may not necessarily have biased the results unless the selection criteria correlate with different underlying cognitive impairments (phonological vs. non-phonological). However, the authors themselves acknowledged this possibility, writing that *"depending on the dyslexia criterion used, samples may vary in composition with regard to their subtypes"* (p.33).

Overall, the presence of artificial admixture was most frequently associated with taxonic conclusions. Only two notable mismatches were identified: @marcus2012latent reached a dimensional conclusion regarding ADHD despite having incurred admixture of clinical cases, while @munson2008asd reached a taxonic conclusion about ASD subtypes without clear evidence of admixture (all children had received an ASD diagnosis based on what appeared to be a uniform clinical assessment). As discussed above, a third exception was observed in a secondary analysis of a subsample by @frazier2010asd targeting ASD. Taken together, these findings reinforce the view that ADHD is best conceptualized as a dimensional condition, while ASD may exhibit taxonic features, although the latter conclusion is supported by limited unbiased evidence.

In our simulated example, we found that artificial admixture represents a major threat to the validity of taxometric analysis, and to clustering methods as well. One might argue that our simulated scenario, in which half of the diagnosed cases were separately sampled, was extreme. However, this design is broadly consistent with the degree of admixture observed in the reviewed literature. It is worth noting that admixture is not limited to the obvious case of oversampling diagnosed individuals, and sometimes occurs in subtle ways. For example, through the undersampling of individuals with low trait levels, or by selectively drawing subsamples based on alternative clinical criteria within a larger dataset, as is often the case in subtype studies [cf. "subtractive" compound sampling, @haslam2020metaanalysis].

An important issue related to the present discussion concerns how we should treat traditional diagnostic categories such as those found in the DSM-5, which have recently come under criticism [@astle2022trandiagnostic; @sonuga2020kuhnian; @sonuga2022dontfit]. We argue that the question of whether clinically relevant conditions reflect discrete categories or multivariate, correlated continua does not necessarily undermine the practical utility of diagnostic systems. While the DSM-5 may seem to endorse a categorical model, it remains atheoretical and does not explicitly or scientifically commit to a taxonic view. Researchers, rather than diagnostic manuals, may risk reifying these categories as natural taxa, thus overinterpreting their conceptual status, hence we advocate for more high-quality taxometric studies.

@@

A first and obvious limitation of the present review concerns the small number of studies identified. This prevented us from conducting a meta-analysis: the available evidence was too scattered, methodologically heterogeneous, and inappropriate for quantitative synthesis. However, this limitation is itself informative, underscoring the need for more taxometric investigation in the field of NDDs. While the search query was comprehensive and no time limits were applied, more studies might have been included by relaxing the eligibility criteria. However, this is unlikely to have made a substantial difference. Exclusion criteria specified the lack of use of behavioral indicators, but no study was excluded on this basis. One record was screened out for being a dissertation [@clemons2006taxometric]. The full document could not be retrieved. Based on its abstract, it is clear that it focused on whether ADHD subtypes represent distinct taxa, and it reached a taxonic conclusion. However, some degree of admixture was possible (participants were separately recruited from clinical centers and schools).

A relevant future direction for this line of investigation concerns a systematic examination of the use and results of cluster analysis. As described above, some of the reviewed articles employed clustering methods, and clustering is a prominent tool in this field [@astle2022trandiagnostic]. We found that model-based approaches, such as latent class/profile analysis, were the most frequently used. The use of clustering is not without problems. First, while it is more flexible and can address broader questions, it does not offer a direct or consistent test of taxonicity [@beauchaine2003taxometrics; @ruscio2004boundaryissues]. Second, meaningful scientific inference from clustering (beyond its simple use as a data reduction technique) requires careful handling of assumptions, as the risk of detecting pseudoclusters is high [@toffalini2022entia; @toffalini2024clusters]. Third, it is is not immune to risks such as artificial admixture. Nonetheless, clustering and taxometric analysis could be more directly, but also critically, examined and compared when addressing the question of whether neurodevelopmental conditions reflect categories or dimensions.

>>>>

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to the topic, and to some of the reviewed articles."
    data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "Taxometric analysis is explicitly designed to test whether the latent structures underlying clinical psychological constructs are categorical (taxonic) or dimensional. Despite its relevance to ongoing debates about the nature of neurodevelopmental disorders (NDDs), its application in this field has been limited. We conducted a preregistered systematic review to critically evaluate the literature, with a focus on methodological threats such as artificial admixture. All published taxometric studies on autism, ADHD, learning disorders, and language impairment were identified and assessed. Studies on ADHD and language impairment generally supported dimensional structures, whereas most studies on autism and NDD subtypes predominantly reported taxonic findings. However, artificial admixture was frequently observed and undermines the credibility of many taxonic conclusions. To illustrate this risk, we conducted a Monte Carlo simulation showing how admixture can spuriously generate pseudotaxonic results. We urge caution in interpreting the outcomes, and advocate for more high-quality taxometric studies in this area."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

@@

In recent years, research on neurodevelopmental disorders (NDDs) has undergone significant reconsideration. Traditional diagnostic categories defined in the DSM-5 [@dsm5] have increasingly been questioned in favor of a dimensional and transdiagnostic perspective [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial]. Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder \[ASD\] in place of distinct diagnoses). Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

Determining whether NDDs represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial]. More fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.

A dimensional approach aligns well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial]. It also seems well-suited for NDDs, given a growing body of evidence supporting polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome]. Most behavioral traits are understood to result not from one or a few causes, but from the additive effects of many small influences [@chabris2015fourth]. According to the central limit theorem [@fischer2011central], when many individual causes (e.g., discrete genetic variants) contribute additively, the resulting trait tends to approximate a continuous, Gaussian distribution. Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the proposed emergence of data-driven taxa [@astle2022trandiagnostic].

Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues and refined over decades [@meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure. It has been widely applied in psychopathology, and typically suggested that most mental health conditions (e.g., depression, anxiety) are better conceptualized as dimensional [@haslam2012review; @haslam2020metaanalysis]. In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to adjudicate between competing latent categorical vs dimensional models [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric]. These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide quantitative measures to facilitate inference [@ruscio2018using].

Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, with most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis]. In their debate on whether ASD is best conceptualized as a category or a continuum, @happe2021both and @chown2021bit do not even mention such methods. In their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but predominantly emphasize studies using cluster analysis. However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues]. Recent simulations using psychologically realistic data (featuring both some skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters]. Indeed, @astle2022trandiagnostic interpret clustering results as evidence that data-driven groups do not align with DSM-5 categories, rather than as compelling tests of latent structural form.

@@

Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition (i.e., the putative taxon and complement). Ideally, Cohen’s *d* for each indicator should substantially exceed 1 (*d* $\gg$ 1). Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership, indicating redundancy. Low base rate means that too few individuals in the sample belong to the taxon, limiting power (samples with *n* \< 30 for the taxonic group are usually deemed insufficient). This is particularly relevant when studying rare conditions. Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate. For instance, clinical cases may be recruited separately from controls and merged into a single sample. In general, poor indicator validity and low base rate reduce power to detect true taxa, while high nuisance covariance and artificial admixture increase the risk of detecting pseudotaxa by mimicking a categorical structure.

@@

The primary aim of this systematic review is to critically examine the use of taxometric methods in studies of NDDs within the published literature. We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), Specific Learning Disorders (SLDs), and the closely related condition of Language Disorder/Impairment. (The latter was not included in the preregistered protocol and constitutes a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.

A crucial goal was to assess the methodological quality and credibility of taxometric findings. We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and skewed indicator distributions. (Strong skewness can create the illusion of a categorical structure, and it might be relevant when assessing rare or extreme behaviors.)

Following the review, we report a Monte Carlo simulation study designed to empirically illustrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data.

@@

Inclusion criteria were as follows:\
1. The study applied at least one taxometric procedure;\
2. The target population consisted of individuals with NDDs or related traits;\
3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) to infer such a conclusion.

Exclusion criteria were:\
1. The article was theoretical, methodological, or a simulation study without empirical data;\
2. The study did not report any behavioral data (in practice, no screened study was excluded for this criterion);\
3. The study was a review and/or meta-analysis;\
4. The study employed latent structure methods other than taxometric analysis (e.g., only clustering), with no taxometric procedures reported.

In addition, following the preregistered protocol, the search was limited to published, peer-reviewed journal articles. This decision reflected the aim of providing a critical review of the existing empirical literature. Only one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Limitation* section.

@@

Two independent human coders (MC and RP) screened and coded all articles. No AI tools conducted any screening or coding, although coders occasionally used them to help clarify specific article content. All discrepancies were resolved through discussion and were later reviewed by the PI (ET).

@@

```{r}
#| label: tbl-summary
#| tbl-cap: Overview summary of the reviewed articles.
summary_tab = read_excel("Tables/Review-Summary.xlsx",sheet=2)
kable(summary_tab, format = "latex", booktabs = TRUE) |>
  kable_styling(font_size = 8, latex_options = c("striped", "scale_down")) |>
  column_spec(1:ncol(summary_tab),width="5em") |>
  landscape()
```

@@

# A Monte Carlo Simulation of Artificial Admixture

Artificial admixture represents a powerful threat to validity that is common but also easily avoidable by design. Different types of admixture can be incurred. @ruscio2004checklist describe three cases. First, when clinical and nonclinical participants are separately recruited and merged into a single sample. Second, when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf. "subtractive" compound sampling, @haslam2020metaanalysis]. Third, when researchers trim observations from the putative complement (individuals without a diagnosis) to inflate the base rate of the taxon. Each of these procedures risks imposing a categorical structure onto data that might be dimensional in nature. Importantly, admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the exact same reason.

@@

We simulate a dimensional scenario to visually demonstrate how artificial admixture, specifically of the first type described above, can produce pseudotaxonic results. A single normally distributed latent variable $X$ is generated to represent the underlying clinical trait of interest. A set of observed indicators ($x_1$, $x_2$, $x3$) are then simulated as linear functions of $X$ plus normally distributed random noise, yielding an $R^2$ of 0.77 for each indicator. Individuals above the 99th percentile of $X$ are assumed to receive a diagnosis via global clinical assessment. A sample of $N = 1000$ individuals is drawn from the general population. Given the low base rate, only about $n = 10$ cases are expected to naturally receive a diagnosis. To address this, researchers draw a second sample consisting of an additional $n = 1000$, targeting clinically diagnosed cases, and merge it with the original sample, thereby creating an artificial admixture. The full simulation code and data are available in the online materials:\
<https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba>

@@

Concerning subtypes, the situation was similar. All four reviewed articles reached some taxonic conclusions, but three of them likely involved some form of artificial admixture. However, the issue appeared to be nuanced and less severe in this subset of studies. For example, @obrien2012dyslexia used psychometric cutoffs across different tools and alternate inclusion criteria, which may not necessarily have biased the results, unless the selection criteria correlate with different underlying cognitive impairments in dyslexia (phonological vs. non-phonological). However, the authors themselves acknowledged this possibility, writing that *"depending on the dyslexia criterion used, samples may vary in composition with regard to their subtypes"* (p.33).

Overall, the presence of artificial admixture was most frequently associated with taxonic conclusions. Only two notable mismatches were identified: @marcus2012latent reached a dimensional conclusion regarding ADHD despite having incurred admixture of clinical cases, while @munson2008asd reached a taxonic conclusion about ASD subtypes without evidence of admixture (all children had received an ASD diagnosis based on a seemingly uniform clinical assessment). Together with the above discussed exception in a secondary analysis of a subsample by @frazier2010asd targeting ASD, these findings reinforce the view that ADHD is best conceptualized as dimensional, while ASD may exhibit taxonic features. However, the latter conclusion is supported by limited unbiased evidence.

In our simulated example, we found that artificial admixture represents a major threat to the validity of taxometric analysis, and to clustering methods as well. One might argue that our simulated scenario, in which half of sample was separately recruited, was extreme. However, this design is broadly consistent with the degree of admixture observed in the reviewed literature. It is worth noting that admixture is not limited to the obvious case of oversampling diagnosed individuals, and sometimes occurs in subtle ways. For example, by selectively drawing subsamples based on alternative clinical criteria within a larger dataset, as is often the case in subtype studies [see "subtractive" compound sampling, @haslam2020metaanalysis].

An important issue related to the present discussion concerns how we should treat traditional diagnostic categories such as those found in the DSM-5, which have recently come under criticism [@astle2022trandiagnostic; @sonuga2020kuhnian; @sonuga2022dontfit]. We argue that the question of whether clinically relevant conditions reflect discrete categories or multivariate, correlated continua does not necessarily undermine the practical utility of diagnostic systems. While the DSM-5 may seem to endorse a categorical model, it remains atheoretical and does not explicitly or scientifically commit to a taxonic view. Researchers, rather than diagnostic manuals, may risk reifying these categories as natural taxa, overinterpreting their conceptual status. Hence we advocate for more high-quality taxometric studies.

@@

A first and obvious limitation of the present review concerns the small number of studies identified, and the virtual lack of studies targeting learning disorders as taxa. This prevented us from conducting a meta-analysis: the available evidence was too scattered and methodologically heterogeneous. However, this limitation is itself informative, underscoring the need for more taxometric investigation in the field. More studies might have been included by relaxing the eligibility criteria. However, this is unlikely to have made a substantial difference. Exclusion criteria specified the lack of use of behavioral indicators, but no study was excluded on this basis. One record was screened out for being a dissertation [@clemons2006taxometric]. The full document could not be retrieved. Based on its abstract, it is clear that it focused on whether ADHD subtypes represent distinct taxa, and it reached a taxonic conclusion. However, some degree of admixture was possible (participants were separately recruited from clinical centers and schools).

A relevant future direction for this line of investigation concerns a systematic examination of the use and results of cluster analysis. Indeed, some of the reviewed articles employed clustering methods, and clustering is a prominent tool in this field [@astle2022trandiagnostic]. Model-based approaches, such as latent class/profile analysis, were the most frequently used. The use of clustering is not without problems. First, while it is more flexible and can address broader questions, it does not offer a direct or consistent test of taxonicity [@beauchaine2003taxometrics; @ruscio2004boundaryissues]. Second, meaningful scientific inference from clustering (beyond its simple use as a data reduction technique) requires careful handling of assumptions, as the risk of detecting pseudoclusters is high [@toffalini2022entia; @toffalini2024clusters]. Finally, it is is not immune to risks such as artificial admixture. Nonetheless, clustering and taxometric analysis could be more directly, but also critically, compared when addressing the question of whether neurodevelopmental conditions reflect categories or dimensions.
